\section{Build systems, concretely}\label{sec-implementations}\label{sec-design-space}

\begin{table}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-2mm}
% \smaller
\centering
\begin{tabular}{lccc}
\hline
 & \multicolumn{3}{c}{\textbf{Scheduling algorithm}}\vspace{1.5mm} \\
\textbf{Rebuilding strategy}\gap  & \gap{}Topological\gap\S\ref{sec-topological}\gap & \gap{}Restarting\gap\S\ref{sec-restarting}\gap & \gap{}Suspending\gap\S\ref{sec-suspending}\gap    \\\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                           & \Make       & \Excel & -              \\
Verifying traces\hfill\S\ref{sec-verifying-traces}                             & \Ninja      & -      & \Shake         \\
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}           & \CloudBuild & \Bazel & -              \\
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{-2mm}
\end{table}

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler (\S\ref{sec-scheduler})
and a rebuilder (\S\ref{sec-rebuilder}).
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder. The result can be summarized in Table~\ref{tab-build-systems},
which tabulates combinations of the scheduling algorithm and the rebuilding
strategy, providing 12~possible build systems, 8~of which are inhabited
by existing build systems (we discuss these systems in~\S\ref{sec-background} and
\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table is suspending
constructive traces, which would provide many benefits, and which we title
\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
% (as we plan on extending \Shake to occupy that spot)

\subsection{Concrete Implementations}

We can define schedulers and rebuilders more concretely with the types
(Fig.~\ref{fig-types}):

\vspace{1mm}
\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
\end{minted}
\vspace{1mm}

\noindent
A \hs{Scheduler} is a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes use of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

A \hs{Rebuilder} takes three arguments: a key, its current value, and a
\hs{Task} that can (re)compute the value of the key if necessary. It uses the
persistent build information \hs{ir} (carried by the state monad) to decide
whether to rebuild the value. If doing so is unnecessary, it returns the current
value; otherwise it runs the supplied \hs{Task} to rebuild it. In both cases it
can choose to update the persistent build information \hs{ir} to reflect what
happened. So a \hs{Rebuilder} wraps a \hs{Task}~\hs{c}~\hs{k}~\hs{v}, which
unconditionally rebuilds the key, to make a
\hs{Task}~\hs{(MonadState}~\hs{ir)}~\hs{k}~\hs{v}, which rebuilds the key only
if necessary, and does the necessary book-keeping. Note that the resulting
\hs{Task} is always monadic; static dependency analysis can be performed on the
original \hs{Task}~\hs{Applicative} if needed.

The scheduler calls the rebuilder, but passes it a \hs{fetch} function
that the latter calls when it needs the value of a dependent key.  This
callback returns control to the scheduler, which may in turn call the
rebuilder to bring the dependent key up to date, and so on.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ mapM_ build order
  where
    build :: k -> State (Store i k v) ()
    build key = case tasks key of
        Nothing -> return ()
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState i) k v
                newTask = rebuilder key value task
                fetch :: k -> State i v
                fetch k = return (getValue k store)
            newValue <- liftStore (run newTask fetch)
            modify $ putValue key newValue
    order = topSort (reachable dep target)
    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do
    (a, newInfo) <- gets (runState x . getInfo)
    modify (putInfo newInfo)
    return a
\end{minted}
% \vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
% \vspace{-5mm}
\end{figure}

These two abstractions are the key to modularity: \emph{we can combine any
scheduler with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of
Table~\ref{tab-build-systems}, and a rebuilder for each row; then combine them
to obtain the build systems in the table's body.

\subsection{\Make}\label{sec-implementation-make}

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. As promised, its definition
is just the application of a \hs{Scheduler}, \hs{topological},
to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
We discuss each component in turn, starting with the rebuilder.

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a state monad. This \hs{MakeInfo} comprises
the \emph{current time} \hs{now}~\hs{::}~\hs{Time} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file modification
times}. We assume that the external system, which invokes the build system,
updates \hs{MakeInfo} reflecting any file changes between successive builds.

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make relies on the
file system to track file modification times, but we prefer to make this
explicit in our model.}; otherwise we can just return the current \hs{value}.

\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based
on a topological sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue}. Note that \hs{newTask} has access only to the \hs{i} part of the
\hs{Store}~\hs{i}~\hs{k}~\hs{v}, but the rest of the \hs{do} block runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad; we use the (unremarkable)
helper function \hs{liftStore} to fix the mismatch. The \hs{newTask} finds
values of the \hs{key}'s dependencies via the \hs{fetch} callback, which is
defined to directly read the \hs{store}.

The pre-processing stage uses the function \hs{dependencies}, defined
in~\S\ref{sec-deps}, to extract static dependencies from the provided
applicative \hs{task}. We compute the linear processing \hs{order} by
constructing the graph of keys \hs{reachable} from the \hs{target} via
dependencies, and performing the topological sort of the result. We omit
implementation of textbook graph algorithms \hs{reachable} and \hs{topSort},
e.g. see~Cormen~\etal~\shortcite{cormen2001introduction}.

Note that the function \hs{dependencies} can only be applied to applicative
tasks, which restricts \Make to static dependencies, as reflected in the
type~\hs{Build}~\hs{Applicative}. Moreover, any other build system that uses
the \hs{topological} scheduler will also inherit the same restriction.

\subsection{\Excel}\label{sec-implementation-excel}

Our model of \Excel uses the \hs{restarting} scheduler and the
\hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
\hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
calc chain of type \hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

The external system, which invokes \Excel's build engine, is required
to provide a transitively closed set of dirty bits. That is, if a
cell is changed, its dirty bit is set, as well as the dirty bit of any other
cell whose value might perhaps change as a result. It is OK to mark too many
cells as dirty; but not OK to mark too few.

The \hs{dirtyBitRebuilder} is very simple: if the \hs{key}'s dirty bit is set,
we \hs{run} the \hs{task} to rebuild the \hs{key}; otherwise we return the
current \hs{value} as is.
Because the dirty cells are transitively closed,
unlike \Make's \hs{modTimeRebuilder}, the \hs{dirtyBitRebuilder} does
not need to modify \hs{i} to trigger rebuilds of dependent keys.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty
                          $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
      Nothing -> (key :) <$> go (Set.insert key done) keys
      Just task -> do
        store <- get
        let newTask :: Task (MonadState ir) k (Either k v)
            newTask = try $ rebuilder key (getValue key store) task
            fetch :: k -> State ir (Either k v)
            fetch k | k `Set.member` done = return $ Right (getValue k store)
                    | otherwise           = return $ Left k
        result <- liftStore (run newTask fetch) -- liftStore is in Fig. 8
        case result of
            Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
            Right newValue -> do modify $ putValue key newValue
                                 (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
% \vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
% \vspace{-2.5mm}
\end{figure}

\Excel's \hs{restarting} scheduler processes keys in the order specified by the
calc \hs{chain}. During the build, it constructs a \hs{newChain} for the next
build and maintains a set of keys \hs{done} that have been processed. For each
non-input \hs{key}, the scheduler tries to rebuild it using a partial \hs{fetch}
callback that returns \hs{Either}~\hs{k}~\hs{v} instead of \hs{v}. The callback
is defined to fail with \hs{Left}~\hs{dep} when asked for the value of a
dependency \hs{dep} that has not yet been processed (and hence may potentially
be dirty); otherwise it returns the current value of the dependency by looking
it up in the \hs{store}.

After the \hs{newTask} is executed (with the help of \hs{liftStore}) there are
two cases to consider:

\begin{itemize}
    \item The \hs{newTask} has failed, because one of its dependencies \hs{dep}
    has not yet been processed. This indicates that the calculation \hs{chain}
    from the previous build is incorrect and needs to be adjusted by moving the
    \hs{dep} in front of the \hs{key}, so that we can restart building the
    \hs{key} after the \hs{dep} is ready.
    \item The \hs{newTask} succeeded. The resulting \hs{newValue} is written to
    the store, the \hs{key} is marked as \hs{done}, and \Excel continues to
    build the rest of the \hs{chain}.
\end{itemize}

Note that the task returned by the \hs{rebuilder} expects a total callback
function and cannot be directly executed with the partial callback \hs{fetch}.
We fix the mismatch with the function \hs{try} that relies on the standard
monad transformer \hs{ExceptT} from the \cmd{transformers} library. We also
need the helper \hs{liftChain}, whose implementation we omit since it
is analogous to \hs{liftStore} in Fig.~\ref{fig-make-implementation}.

\subsection{\Shake}\label{sec-implementation-shake}

Our model of \Shake (Fig.~\ref{fig-shake-implementation}) stores verifying
traces \hs{VT}~\hs{k}~\hs{v} defined in~\S\ref{sec-verifying-traces} as
persistent build information and is composed of the \hs{suspending} scheduler
and the \hs{vtRebuilder}.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store =
    fst $ execState (fetch target) (store, Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
% \vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
% \vspace{-3mm}
\end{figure}

The rebuilder performs the \hs{verifyVT} query to determine if the \hs{key} is
\hs{upToDate}. If it is, the rebuilder simply returns the \hs{key}'s current
\hs{value}. Otherwise it executes the \hs{task}, obtaining both a \hs{newValue}
and the \hs{key}'s dynamic dependencies \hs{deps} (see the definition of
\hs{track} in~\S\ref{sec-deps}), which are subsequently recorded in the trace
store using \hs{recordVT}.

The \hs{suspending} scheduler uses a recursive \hs{fetch} callback, defined
similarly to the \hs{busy} build system (\S\ref{sec-general-build}), that builds
a given \hs{key}, making sure not to duplicate work when called on the same
\hs{key} again in future. To achieve that, it keeps track of keys that have
already been built in a set \hs{done}~\hs{::}~\hs{Set}~\hs{k}. Given a non-input
\hs{key} that has not yet been built, we use the supplied \hs{rebuilder} to
embed the build information~\hs{i} into the \hs{task}. We then execute the
obtained \hs{newTask} by passing it the \hs{fetch} function as a callback for
building dependencies: the \hs{newTask} will therefore be suspended while its
dependencies are being brought up to date. The \hs{newValue} obtained by running
the \hs{newTask} is stored, and the \hs{key} is added to the set \hs{done}.

The \hs{fetch} computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v,}~\hs{Set}~\hs{k)} monad. To make
\hs{MonadState}~\hs{i} access the \hs{i} inside the \hs{Store} we use the helper
function \hs{liftRun} (which uses a \hs{newtype} to provide a \hs{MonadState}
instance that sees through into the \hs{Store}).

As discussed in \S\ref{sec-step-traces}, \Shake actually use verifying step traces, but we
choose to focus on the more explicit verifying traces. We have implemented
verifying step traces in our framework, and they compose with schedulers as
you would hope.

\subsection{Cloud Build Systems: \Bazel, \CloudBuild, \Cloud \Shake, \Buck and \Nix}
\label{sec-implementation-cloud}

Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
variant (\S\ref{sec-deep-constructive-traces}).

The implementation of \hs{ctRebuilder} is analogous to that of \hs{vtRebuilder}
in Fig.~\ref{fig-shake-implementation}, but the \hs{verifyVT} query is replaced
with a more powerful query to \hs{constructCT} that returns a list of suitable
\hs{cachedValues} by looking them up the cloud cache. If the current \hs{value}
is in the list, we can use it as is. Otherwise, if the list is non-empty, we can
use an arbitrary \hs{cachedValue}. Finally, if the cache has no suitable values,
we fall back to executing the \hs{task}. The obtained \hs{newValue} and the
\hs{task}'s dependencies are recorded as a new constructive trace for future
use.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restartingQ ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting scheduler based on a build queue, omitted (22 lines)
restartingQ :: (Hashable v, Eq k) => Scheduler Monad (CT k v) (CT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    case cachedValues of
        _ | value `elem` cachedValues -> return value
        cachedValue:_ -> return cachedValue
        [] -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 10
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 8
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 8
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to 'ctRebuilder'
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 10
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
% \vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix in our framework.}
\label{fig-cloud-implementations}
% \vspace{-2mm}
\end{figure}

The \Bazel build system uses a restarting scheduler whose implementation we
omit. It is similar to \Excel's \hs{restarting} scheduler defined in
Fig.~\ref{fig-excel-implementation}, but instead of building keys in the order
specified by the persistently stored calc chain, \Bazel uses a \emph{build
queue}. The build starts with the queue containing all dirty keys. Similar to
\Excel, the rebuilding of a key extracted from the queue may fail because one of
its dynamic dependencies is dirty. In this case the key is marked as
\emph{blocked} and its rebuilding is deferred. Whenever a key is successfully
rebuilt, all keys that were previously blocked on it are added back to the
queue, and their build is eventually restarted.

Note that although both our model and \Bazel's actual implementation supports
dynamic dependencies, it is currently not possible to define new monadic build
rules in the language available to users. Instead, users have to rely on a
collection of predefined built-in rules, which cover many
important instances of dynamic dependencies.

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs of
dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
we need to convert a monadic \hs{ctRebuilder} into an applicative one by
applying an adapter function \hs{adaptRebuilder}, which unwraps a given
\hs{Task}~\hs{Applicative} and wraps it into \hs{Task}~\hs{Monad}.

Our models of \Buck~\cite{buck} and \Nix~\cite{dolstra2004nix} use the rebuilder
based on deep constructive traces (\S\ref{sec-deep-constructive-traces}), called
\hs{dctRebuilder}, whose implementation we omit since it is very similar to that
of \hs{ctRebuilder}. \Buck uses the \hs{topological} scheduler and is therefore
an applicative build system, whereas \Nix uses the \hs{suspending} scheduler and
is monadic.

\todo{AM}{Add implementation of \hs{dctRebuilder}?}

Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is combine \hs{suspending} with \hs{ctRebuilder},
as shown in Fig.~\ref{fig-cloud-implementations}.
